Davos: a Python package ``smuggler'' for constructing lightweight reproducible notebooks

### 2023-09-17 paper/main.tex
additional text in projects section

For greater control over the behavior of \texttt{smuggle} statements, Davos
defines an additional construct called the ``onion comment.'' An onion comment
is a special type of inline comment that may be placed on a line containing a
\texttt{smuggle} statement to customize how Davos searches for the smuggled
package locally and, if necessary, downloads and installs it. Onion comments
follow a simple format based on the ``type comment'' syntax introduced in PEP
484~\cite{vanREtal14}, and are designed to make managing packages with Davos
intuitive and familiar. To construct an onion comment, users provide the name
of the installer program (e.g., \texttt{pip}) and the same arguments one would
use to manually install the package as desired via the command line:
\begin{center}
  \includegraphics[width=0.9\textwidth]{figs/snippet1}
\end{center}
Occasionally, a package's distribution name (i.e., the name used
when installing it) may differ from its top-level module name (i.e., the name
used when importing it). In such cases, an onion comment may be used to ensure
that Davos installs the proper package if it cannot be found locally:
\begin{center}
  \includegraphics[width=0.9\textwidth]{figs/snippet2}
\end{center}
Because onion comments may be constructed to specify any aspect of
the installer program's behavior, they provide a mechanism for precisely controlling
how, where, and when smuggled packages are installed. Critically, if an onion
comment includes a version specifier~\cite{CoghStuf13}, Davos will ensure that
the version of the package loaded into the notebook matches the specific
version requested (or satisfies the given version constraints). If the smuggled
package exists locally, Davos will extract its version information from its
metadata and compare it to the specifier provided. If the two are incompatible
(or no local installation is found), Davos will download, install, and load a
suitable version of the package instead:
\begin{center}
  \includegraphics[width=0.9\textwidth]{figs/snippet3}
\end{center}
Onion comments can also be used to \texttt{smuggle} specific VCS references (e.g.,
Git~\cite{TorvHama05} branches, commits, tags, etc.):
\begin{center}
  \includegraphics[width=0.9\textwidth]{figs/snippet4}
\end{center}
Davos processes onion comments internally before forwarding arguments to the
installer program. In addition to preventing shared notebooks from executing
arbitrary code in a user's shell, this enables Davos to adjust its behavior
based on how particular flags will affect the behavior of the installer
program. For example, including \texttt{pip}'s \texttt{--no-input} flag will also
temporarily enable Davos's non-interactive mode (see Sec.~\ref{subsec:config}).
Similarly, if an onion comment contains either \texttt{-I}/\texttt{--ignore-installed},
\texttt{-U}/\texttt{--upgrade}, or \texttt{--force-reinstall}, Davos will
install and load a new copy of the smuggled package without first checking
for it locally:
\begin{center}
  \includegraphics[width=0.9\textwidth]{figs/snippet5}
\end{center}
Since the purpose of an onion comment is to describe how a smuggled package
should be installed (if necessary) so that it can be loaded and used
immediately, options that would normally cause the package not to be installed
(such as \texttt{-h}/\texttt{--help} or \texttt{--dry-run}) are disallowed. Additionally,
when using a Davos ``project'' to isolate smuggled packages (the default behavior;
see Sec.~\ref{subsec:projects}), onion comments may not contain options that
would change the package's installation location (such as
\texttt{-t}/\texttt{--target}, \texttt{--root}, or \texttt{--prefix}). However, if
the user disables project-based isolation and specifies \texttt{--target <dir>},
Davos will ensure that \texttt{<dir>} is included in the module search path (i.e.,
\texttt{sys.path}), prepending it if necessary, so the package can be loaded.

@@

To prevent Davos-enhanced notebooks from having unwanted side-effects on the user's environment, packages installed via \texttt{smuggle} statements are automatically isolated using a custom, virtual environment-like system called ``projects.'' 
Davos projects are similar to standard Python virtual environments (e.g., created with the standard library's \texttt{venv} module or a third-party tool like \texttt{virtualenv}~\cite{BickEtal07}) but with a few noteworthy differences that make them generally lighter-weight and simpler to use. 
Like a standard virtual environment, a Davos project consists of a directory (within a hidden \texttt{.davos} folder in the user's home directory) that houses third-party packages needed for a particular Python project, workflow, or task. 
However, unlike standard virtual environments, Davos projects do not need to be manually created, activated, or deactivated, and function to \textit{extend} the user's existing Python environment rather than replace it.

When Davos is imported into a notebook, a project directory for that notebook is automatically created (if it does not exist already). 
When \texttt{smuggle} statements within that notebook are then executed, any packages (or specific versions of packages) that are not already available in the user's Python environment will be installed into the notebook's project directory (along with any missing dependencies of those packages). 
During each \texttt{smuggle} statement's execution, Davos also temporarily prepends the notebook's project directory to the module search path so that these project-installed packages are visible when searching for smuggled packages locally, and prioritized over packages in the user's main environment.

Thus, rather than constructing fully separate Python environments from scratch, Davos projects work by supplementing the user's existing environment with any additional packages (or specific package versions) needed to satisfy the dependencies of their corresponding notebooks.
In some cases, this might include every package smuggled into a notebook (e.g., if the notebook is run inside a freshly created, empty virtual environment). 
In other cases, the user's environment may already contain all required packages, and the notebook's project directory will be unused (in which case it will be deleted automatically when the notebook kernel is shut down). 
But regardless of the extent to which the existing environment is augmented, Davos's project system ensures that all smuggled packages are installed locally and can be loaded successfully at runtime, but that the contents of the user's Python environment are never altered. 
Additionally, because \texttt{smuggle} statements in a given notebook are evaluated every time it is run, this system also ensures that the notebook's requirements will continue to be satisfied even if the user's Python environment changes. 
For example, suppose a user has \texttt{NumPy}~\cite{HarrEtal20} v1.24.3 installed in their current Python environment and runs a Davos-enhanced notebook that smuggles \texttt{NumPy} with ``\texttt{numpy==1.24.3}'' specified in an onion comment (see Sec.~\ref{subsec:onion}). 
Since the user's existing version of the package satisfies this requirement, Davos will happily load it into the notebook. 
But if the user later upgrades their environment's \texttt{NumPy} version to v1.25.0 (perhaps as a result of installing a different package that depends on it) and subsequently reruns the notebook, the local version will longer satisfy the requirement, so Davos will install \texttt{NumPy} v1.24.3 into the notebook's project directory and load that version instead. 
From then on, further changes to the user's \texttt{Numpy} installation would not have any effect on Davos's behavior in this particular notebook, as a satisfactory version now exists in its project directory (if the version in the onion comment were changed, Davos would install it in the project directory in place of the current version). 
For efficiency, Davos projects will generally avoid duplicating dependencies already satisfied by the user's Python environment. 
However, if desired, adding \texttt{pip}'s \texttt{--ignore-installed} flag to an onion comment in the notebook will cause the smuggled package to be installed in the project directory whether or not it already exists locally.

By default, Davos projects are named based on the absolute path to their corresponding notebooks, and are therefore notebook-specific.

>>>>

For greater control over the behavior of \texttt{smuggle} statements, Davos
defines an additional construct called the ``onion comment.'' An onion comment
is a special type of inline comment that may be placed on a line containing a
\texttt{smuggle} statement to customize how Davos searches for the smuggled
package locally and, if necessary, downloads and installs it. Onion comments
follow a simple format based on the ``type comment'' syntax introduced in PEP
484~\cite{vanREtal14}, and are designed to make managing packages with Davos
intuitive and familiar. To construct an onion comment, users provide the name
of the installer program (e.g., \texttt{pip}) and the same arguments one would
use to manually install the package as desired via the command line:
\begin{center}
  \includegraphics[width=0.9\textwidth]{figs/snippet1}
\end{center}
Occasionally, a package's distribution name (i.e., the name used
when installing it) may differ from its top-level module name (i.e., the name
used when importing it). In such cases, an onion comment can be used to ensure
that Davos installs the proper package if it cannot be found locally:
\begin{center}
  \includegraphics[width=0.9\textwidth]{figs/snippet2}
\end{center}
Because onion comments may be constructed to specify any aspect of
the installer program's behavior, they provide a mechanism for precisely controlling
how, where, and when smuggled packages are installed. Critically, if an onion
comment includes a version specifier~\cite{CoghStuf13}, Davos will ensure that
the version of the package loaded into the notebook matches the specific
version requested (or satisfies the given version constraints). If the smuggled
package exists locally, Davos will extract its version information from its
metadata and compare it to the specifier provided. If the two are incompatible
(or no local installation is found), Davos will download, install, and load a
suitable version of the package instead:
\begin{center}
  \includegraphics[width=0.9\textwidth]{figs/snippet3}
\end{center}
Onion comments can also be used to \texttt{smuggle} specific VCS references (e.g.,
Git~\cite{TorvHama05} branches, commits, tags, etc.):
\begin{center}
  \includegraphics[width=0.9\textwidth]{figs/snippet4}
\end{center}
Davos processes onion comments internally before forwarding arguments to the
installer program. In addition to preventing shared notebooks from executing
arbitrary code in a user's shell, this enables Davos to adjust its behavior
based on how particular flags will affect the behavior of the installer
program. For example, including \texttt{pip}'s \texttt{--no-input} flag will also
temporarily enable Davos's non-interactive mode (see Sec.~\ref{subsec:config}).
Similarly, if an onion comment contains either \texttt{-I}/\texttt{--ignore-installed},
\texttt{-U}/\texttt{--upgrade}, or \texttt{--force-reinstall}, Davos will
install and load a new copy of the smuggled package without first checking
for it locally:
\begin{center}
  \includegraphics[width=0.9\textwidth]{figs/snippet5}
\end{center}
Since the purpose of an onion comment is to describe how a smuggled package
should be installed (if necessary) so that it can be loaded and used
immediately, options that would normally cause the package not to be installed
(such as \texttt{-h}/\texttt{--help} or \texttt{--dry-run}) are disallowed. Additionally,
when using a Davos ``project'' to isolate smuggled packages (the default behavior;
see Sec.~\ref{subsec:projects}), onion comments may not contain options that
would change the package's installation location (such as
\texttt{-t}/\texttt{--target}, \texttt{--root}, or \texttt{--prefix}). However, if
the user disables project-based isolation and specifies \texttt{--target <dir>},
Davos will ensure that \texttt{<dir>} is included in the module search path (i.e.,
\texttt{sys.path}), prepending it if necessary, so the package can be loaded.

@@

To prevent Davos-enhanced notebooks from having unwanted side-effects on the user's environment, any packages installed via \texttt{smuggle} statements are automatically isolated using a custom, virtual environment-like system called ``projects.'' 
Davos projects are similar to standard Python virtual environments (e.g., created with the standard library's \texttt{venv} module or a third-party tool like \texttt{virtualenv}~\cite{BickEtal07}) but with a few noteworthy differences that make them generally lighter-weight and simpler to use. 
Like a standard virtual environment, a Davos project consists of a directory (within a hidden \texttt{.davos} folder in the user's home directory) that houses third-party packages needed for a particular Python project, workflow, or task. 
However, unlike standard virtual environments, Davos projects do not need to be manually created, activated, or deactivated, 
%and work by \textit{supplementing} the user's existing Python environment rather than replacing it.
and function to \textit{extend} the user's existing Python environment rather than replace it.

When Davos is imported into a notebook, a project directory for that notebook is automatically created (if it does not exist already). 
When \texttt{smuggle} statements within that notebook are then executed, any packages (or specific versions of packages) that are not already available in the user's Python environment are installed into the notebook's project directory (along with any missing dependencies of those packages). 
During each \texttt{smuggle} statement's execution, Davos also temporarily prepends the notebook's project directory to the module search path so that these project-installed packages are visible when searching for smuggled packages locally, and prioritized over those in the user's main environment.

Thus, rather than constructing fully separate Python environments from scratch, 
%Davos projects serve to fill in the gap between the user's existing environment and the dependencies of their corresponding notebooks. 
Davos projects work by supplementing the user's existing environment with any additional packages (or specific package versions) needed to satisfy the dependencies of their corresponding notebooks.
In some cases, this might include every package smuggled into a notebook (e.g., if the notebook is run inside a freshly created, empty virtual environment). 
In other cases, the user's environment may already provide all required packages, and the notebook's project directory will be unused (in which case it will be deleted automatically when the notebook kernel is shut down). 
But regardless of the extent to which the existing environment is augmented, Davos's project system ensures that all smuggled packages are installed locally and loaded successfully at runtime, while the contents of the user's Python environment are never altered. 
Additionally, because \texttt{smuggle} statements in a given notebook are evaluated every time it is run, this system also ensures that the notebook's requirements will continue to be satisfied even if the user's Python environment changes. 
For example, suppose a user has \texttt{NumPy}~\cite{HarrEtal20} v1.24.3 installed in their current Python environment and runs a Davos-enhanced notebook that smuggles \texttt{NumPy} with ``\texttt{numpy==1.24.3}'' specified in an onion comment (see Sec.~\ref{subsec:onion}). 
Since the user's existing version of the package satisfies this requirement, Davos will happily load it into the notebook. 
But if the user later upgrades their environment's \texttt{NumPy} version to v1.25.0 (perhaps as a result of installing a different package that depends on it) and subsequently reruns this notebook, the local version will longer satisfy this requirement, so Davos will install \texttt{NumPy} v1.24.3 into the notebook's project directory and load that version instead. 
From then on, any further changes to the user's \texttt{Numpy} installation would have no effect on Davos's behavior in this particular notebook, as a satisfactory version now exists in its project directory. (If the version specified in the onion comment were changed, Davos would install it in the project directory in place of the previous version.) 
For efficiency, Davos projects will generally avoid duplicating dependencies already satisfied by the user's Python environment. 
However, if desired, adding \texttt{pip}'s \texttt{--ignore-installed} flag to an onion comment in the notebook will cause the smuggled package to be installed into the project directory whether or not it already exists locally.
